---
title: "Liftoff Atlas | Visuals from JSR Orbital Launch History (1957–2024)"
author: "Dev Verma"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

# Core packages
library(readr)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(janitor)

# Mapping packages
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Create outputs folder (relative to this .Rmd)
if (!dir.exists("outputs")) dir.create("outputs")

# Theme colours for Liftoff Atlas
atlas_pink <- "#FF4DA6"
atlas_grey <- "#9AA0A6"

# Historical bounds
year_min_hist <- 1957L
year_max_hist <- 2024L

# Paths (this .Rmd lives in MavenSpace/, data lives in MavenSpace/data/)
jsr_path   <- "data/jsr_unique_launches_1957_2024.csv"
sites_path <- "data/LaunchSites_Final_Geocoded.csv"
```

##Load canonical JSR dataset & launch-site geography

```{r}
# Load JSR canonical dataset (6,617 launches)

jsr <- read_csv(jsr_path, show_col_types = FALSE) %>%
clean_names() %>%
mutate(
launch_date = as_datetime(dt_min),
year        = year(launch_date)
) %>%
filter(
!is.na(year),
year >= year_min_hist,
year <= year_max_hist
)

nrow(jsr)
n_distinct(jsr$launch_id)

# Load geocoded launch sites

sites <- read_csv(sites_path, show_col_types = FALSE) %>%
clean_names()

head(sites)
```

## 2. Outcome breakdown (Success / Failure / Partial)

```{r}
jsr_kpi <- jsr %>%
mutate(
outcome_group = dplyr::case_when(
success_hard == TRUE                             ~ "Success",
success_hard == FALSE & success == FALSE         ~ "Failure",
TRUE                                             ~ "Partial / other"
)
) %>%
count(outcome_group, name = "launches") %>%
mutate(
total = sum(launches),
share = round(100 * launches / total, 1)
)

jsr_kpi
```

## 3. Humanity’s orbital cadence (1957–2024)
An annual time-series of orbital launches using only the canonical JSR dataset.

```{r}
jsr_yearly <- jsr %>%
count(year, name = "launches") %>%
arrange(year)

head(jsr_yearly)

# Era bands for subtle background context

eras <- tibble::tribble(
~xmin, ~xmax, ~label,
1957, 1975, "Space Race",
1981, 2011, "Space Shuttle Era",
2012, 2024, "NewSpace & Reusability"
)

p_humanity <- ggplot() +

# Era shading

geom_rect(
data  = eras,
aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
fill  = "black",
alpha = 0.03,
inherit.aes = FALSE
) +
geom_text(
data = eras,
aes(x = (xmin + xmax) / 2, y = Inf, label = label),
vjust = 1.4,
size  = 3,
colour = "#666666"
) +

# Area + line

geom_area(
data = jsr_yearly,
aes(x = year, y = launches),
fill  = atlas_pink,
alpha = 0.35
) +
geom_line(
data = jsr_yearly,
aes(x = year, y = launches),
colour    = atlas_pink,
linewidth = 0.9
) +
scale_x_continuous(
breaks = seq(1960, 2020, by = 10),
expand = expansion(mult = c(0, 0.01))
) +
scale_y_continuous(
expand = expansion(mult = c(0, 0.08))
) +
labs(
title    = "Humanity’s Orbital Launches · 1957–2024",
subtitle = "Annual count of verified orbital launches (JSR canonical dataset)",
x        = NULL,
y        = "Launches per year"
) +
theme_minimal() +
theme(
plot.background   = element_rect(fill = "transparent", colour = NA),
panel.background  = element_rect(fill = "transparent", colour = NA),
panel.grid.major.x = element_blank(),
panel.grid.minor   = element_blank(),
legend.position   = "none",
plot.title.position = "plot"
)

p_humanity

```
```{r}
ggsave(
filename = "outputs/line_humanity_jsr_1957_2024.png",
plot     = p_humanity,
width    = 2600 / 300,
height   = 900  / 300,
dpi      = 300,
bg       = "transparent"
)
# The PNG is now ready to drop straight into Figma.
```

## 4. Liftoff Atlas — global launch-site map
Use LaunchSites_Final_Geocoded.csv for the world map bubbles.
This file should already have one row per canonical launch site, with
columns such as site_name, latitude, longitude, and n (launch count).

```{r}
world <- rnaturalearth::ne_countries(
scale       = "medium",
returnclass = "sf"
)

p_atlas <- ggplot() +

# Basemap

geom_sf(
data   = world,
fill   = "black",
colour = "#333333",
linewidth = 0.1
) +

# Launch-site bubbles

geom_point(
data = sites,
aes(
x    = longitude,
y    = latitude,
size = n
),
colour = atlas_pink,
alpha  = 0.45,
stroke = 0
) +
scale_size_area(
max_size = 18,
breaks   = c(50, 250, 500, 1000, 1500),
name     = "Launches"
) +
coord_sf(expand = FALSE) +
labs(
title    = "Liftoff Atlas · Global Orbital Launch Sites",
subtitle = "Bubble size represents total orbital launches per site (JSR 1957–2024)",
x        = NULL,
y        = NULL
) +
theme_void() +
theme(
plot.background   = element_rect(fill = "transparent", colour = NA),
panel.background  = element_rect(fill = "transparent", colour = NA),
legend.position   = "right",
legend.title      = element_text(size = 8),
legend.text       = element_text(size = 7),
plot.title.position = "plot"
)

p_atlas

ggsave(
filename = "outputs/map_liftoff_atlas_jsr.png",
plot     = p_atlas,
width    = 3000 / 300,
height   = 1600 / 300,
dpi      = 300,
bg       = "transparent"
)

```


## 5. SpaceX orbital launches over time
Isolate SpaceX from JSR (agency code "SPX") and show its growth.

```{r}
spacex_ts <- jsr %>%
filter(agency == "SPX") %>%
count(year, name = "launches") %>%
arrange(year)

spacex_ts
```

```{r SpaceX visual}
p_spacex <- ggplot(spacex_ts, aes(x = year, y = launches)) +
geom_line(
colour    = atlas_pink,
linewidth = 1.0
) +
geom_point(
colour = atlas_pink,
size   = 2
) +
scale_x_continuous(
breaks = seq(min(spacex_ts$year, na.rm = TRUE),
max(spacex_ts$year, na.rm = TRUE),
by = 2),
expand = expansion(mult = c(0.01, 0.01))
) +
scale_y_continuous(
expand = expansion(mult = c(0, 0.08))
) +
labs(
title    = "SpaceX Orbital Launches · 2006–2024 (JSR)",
subtitle = "Verified orbital launches attributed to SpaceX in the JSR dataset",
x        = NULL,
y        = "Launches per year"
) +
theme_minimal() +
theme(
plot.background   = element_rect(fill = "transparent", colour = NA),
panel.background  = element_rect(fill = "transparent", colour = NA),
panel.grid.major.x = element_blank(),
panel.grid.minor   = element_blank(),
plot.title.position = "plot"
)

p_spacex

ggsave(
filename = "outputs/line_spacex_jsr_2006_2024.png",
plot     = p_spacex,
width    = 3200 / 300,
height   = 1280 / 300,
dpi      = 300,
bg       = "transparent"
)
```
### You can add milestone labels (first Falcon 1, first Falcon 9, record years) later in Figma.

## 6. Cosmic Swarm A – Launch-true radial visualization (1 dot = 1 launch)
This version uses the canonical JSR launch table:

- 1 row in `jsr_unique_launches_1957_2024.csv` = **1 point**
- Radius encodes **year** (1957 → inner rings, 2024 → outer edge)
- The most active site (Plesetsk) forms the **pink core**
- All other sites form the dark grain field

```{r cosmic_swarm_A, message=FALSE, warning=FALSE}
library(readr);  library(dplyr);  library(stringr); library(janitor)
library(ggplot2); library(ggforce); library(scales)

if (!dir.exists("outputs")) dir.create("outputs")

atlas_pink <- "#FF4DA6"
ink_dark   <- "#000000"
ring_ink   <- "#111111"

set.seed(1957)

csv_jsr   <- "data/jsr_unique_launches_1957_2024.csv"
csv_alias <- "data/site_alias.csv"
csv_canon <- "data/canon_48.csv"

# 1) Load canonical launch table + mapping dictionaries -----------------------
u      <- read_csv(csv_jsr,   show_col_types = FALSE) |> clean_names()
alias  <- read_csv(csv_alias, show_col_types = FALSE) |> clean_names()
canon  <- read_csv(csv_canon, show_col_types = FALSE) |> clean_names()

# Normalise site strings (same logic as data-engineering pipeline)
norm <- function(x){
  x |>
    stringr::str_to_upper() |>
    stringr::str_replace_all("[\u00A0]", " ") |>
    stringr::str_replace_all("[^A-Z0-9 /-]", " ") |>
    stringr::str_squish()
}

u_map <- u |>
  transmute(
    launch_id,
    year  = as.integer(year),
    alias = norm(coalesce(site_raw, pad_raw))
  ) |>
  left_join(alias |> mutate(alias = norm(alias)), by = "alias") |>
  rename(site_key = site_key)

# Hard guardrail: every launch must map to a canonical site
stopifnot(sum(is.na(u_map$site_key)) == 0)

# Identify nucleus site (most launches)
top_site <- u_map |>
  count(site_key, sort = TRUE) |>
  slice(1) |>
  pull(site_key)

# 2) Build 1:1 coordinates (6,617 points) ------------------------------------
year_min <- 1957L
year_max <- 2024L

core_sd  <- 0.028  # size of pink nucleus cloud
ring_min <- 0.10   # inner padding radius
ring_max <- 1.00   # outer radius
j_rad_sd <- 0.006  # small radial jitter

core_df <- u_map |> filter(site_key == top_site)
oth_df  <- u_map |> filter(site_key != top_site)

# Pink core around origin
n_core <- nrow(core_df)
core_xy <- tibble::tibble(
  x   = rnorm(n_core, mean = 0, sd = core_sd),
  y   = rnorm(n_core, mean = 0, sd = core_sd),
  grp = "core"
)

# Map year → radius, then add jitter
scale_r <- function(y){
  t <- (pmax(pmin(y, year_max), year_min) - year_min) / (year_max - year_min)
  rescale(t, to = c(ring_min, ring_max))
}

n_oth <- nrow(oth_df)
ang   <- runif(n_oth, 0, 2 * pi)
rad   <- scale_r(oth_df$year) + rnorm(n_oth, 0, j_rad_sd)

oth_xy <- tibble::tibble(
  x   = rad * cos(ang),
  y   = rad * sin(ang),
  grp = "others"
)

swarm <- dplyr::bind_rows(core_xy, oth_xy)

# Sanity check: 1 dot per launch
nrow(swarm); dplyr::n_distinct(u_map$launch_id)

# Decade guide rings (1957, 1960, …, 2020)
decades <- tibble::tibble(
  year_break = c(1957, seq(1960, 2020, by = 10))
) |>
  mutate(r = scale_r(year_break))

# 3) Plot + export ------------------------------------------------------------
p_cosmic <- ggplot() +
  # Decade rings
  geom_circle(
    data = decades,
    aes(x0 = 0, y0 = 0, r = r),
    color     = alpha(ring_ink, 0.32),
    linetype  = "dotted",
    linewidth = 0.35
  ) +
  # All other launches (dark dust)
  geom_point(
    data  = dplyr::filter(swarm, grp == "others"),
    aes(x, y),
    color = alpha(ink_dark, 0.70),
    size  = 0.28,
    stroke = 0
  ) +
  # Nucleus (Plesetsk) in pink
  geom_point(
    data  = dplyr::filter(swarm, grp == "core"),
    aes(x, y),
    color = atlas_pink,
    alpha = 0.70,
    size  = 0.32,
    stroke = 0
  ) +
  coord_equal(xlim = c(-1.06, 1.06), ylim = c(-1.06, 1.06), expand = FALSE) +
  theme_void() +
  theme(
    plot.background  = element_rect(fill = "transparent", colour = NA),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.margin      = margin(0, 0, 0, 0)
  )

p_cosmic

ggsave(
  filename = "outputs/swarm_historical.png",
  plot     = p_cosmic,
  width    = 4096,
  height   = 4096,
  units    = "px",
  dpi      = 320,
  bg       = "transparent"
)
```

7. Cosmic Swarm B – Dense poster variant (grain halo)

This version does not change the data story.
It builds a synthetic grain field around a pink core to create the dark, inky halo used in the printed poster:

Used only for the final aesthetic in Figma

Launch-true story still comes from Variant A

Exported as a separate PNG so viewers can distinguish them

```{r}
library(dplyr);  library(ggplot2); library(tibble); library(scales)

if (!dir.exists("outputs")) dir.create("outputs")

atlas_pink <- "#FF4DA6"

# --- Parameters (tuned to match poster aesthetic) ---------------------------

set.seed(1957)

# Core (pink nucleus)

N_core   <- 52000      # number of pink points
r_core   <- 0.50       # core radius
j_core   <- 0.06       # small jitter to avoid banding

# Concentric black rings (outer grain halo)

rings <- tibble(
r   = c(0.72, 0.98, 1.25, 1.52, 1.78, 2.00),
sd  = c(0.045, 0.055, 0.060, 0.065, 0.070, 0.080),
pct = c(0.22, 0.24, 0.22, 0.17, 0.10, 0.05)
)

N_total_others <- 260000  # total black dots across all rings

# --- Build pink core (uniform disk + fine jitter) ---------------------------

u_core  <- sqrt(runif(N_core, 0, r_core^2))
th_core <- runif(N_core, 0, 2 * pi)

core <- tibble(
grp   = "core",
x     = u_core * cos(th_core) + rnorm(N_core, 0, j_core),
y     = u_core * sin(th_core) + rnorm(N_core, 0, j_core),
alpha = 0.40,
size  = 0.18
)

# --- Build concentric rings (outer halo) ------------------------------------

set.seed(1957 + 1)

others <- rings |>
rowwise() |>
do({
n  <- round(N_total_others * .$pct)
ph <- runif(n, 0, 2 * pi)
rd <- pmax(r_core + 0.04, rnorm(n, .$r, .$sd))  # keep outside pink core

tibble(
  grp   = "others",
  x     = rd * cos(ph),
  y     = rd * sin(ph),
  # opacity peaks near ring centre → inky bands
  alpha = rescale(abs(rd - .$r), to = c(0.55, 0.15), from = c(.$sd, 0)),
  size  = 0.16
)


}) |>
bind_rows() |>
mutate(alpha = pmin(pmax(alpha, 0.12), 0.60))      # clamp alpha range

swarm_dense <- bind_rows(core, others)

# --- Plot dense swarm -------------------------------------------------------

p_swarm_dense <- ggplot(swarm_dense, aes(x, y)) +
geom_point(
data  = subset(swarm_dense, grp == "others"),
aes(alpha = alpha),
colour = "black",
size   = 0.16,
show.legend = FALSE
) +
geom_point(
data  = subset(swarm_dense, grp == "core"),
colour = atlas_pink,
alpha  = 0.38,
size   = 0.18,
show.legend = FALSE
) +

# subtle bloom at the centre to help the core pop on a light background

annotate("point", x = 0, y = 0, size = 90,
colour = atlas_pink, alpha = 0.10) +
coord_equal(xlim = c(-2.15, 2.15), ylim = c(-2.15, 2.15), expand = FALSE) +
theme_void() +
theme(
plot.background  = element_rect(fill = "transparent", colour = NA),
panel.background = element_rect(fill = "transparent", colour = NA),
plot.margin      = margin(0, 0, 0, 0)
)

p_swarm_dense

ggsave(
filename = "outputs/swarm_dense_poster_github.png",
plot     = p_swarm_dense,
width    = 6000 / 300,
height   = 6000 / 300,
units    = "in",
dpi      = 300,
bg       = "transparent"
)

```



## 8. How these visuals connect to the poster

This document is the visual companion to the data-engineering pipeline.  
Every chart on the *Liftoff Atlas* poster is generated here from the canonical JSR launch table:

- **Humanity’s cadence** (`outputs/line_humanity_jsr_1957_2024.png`)  
  One line per year, using only `jsr_unique_launches_1957_2024.csv`.

- **Liftoff Atlas map** (`outputs/map_liftoff_atlas_jsr.png`)  
  Bubbles are aggregated from `LaunchSites_Final_Geocoded.csv`. Each point is one geocoded launch site with a verified count.

- **SpaceX timeline** (`outputs/line_spacex_jsr_2006_2024.png`)  
  Filters the same table down to `agency == "SPX"` and counts launches per year.

- **Cosmic Swarm A – launch-true** (`outputs/swarm_historical.png`)  
  One point per verified orbital launch. Radius encodes year, colour encodes the most active site versus all others.

- **Cosmic Swarm B – dense poster variant** (`outputs/swarm_dense_poster_github.png`)  
  A synthetic grain field wrapped around a pink core. It keeps the same visual metaphor as A, but adds extra dots to create the inky halo used in the printed poster.  
  The underlying statistics and launch counts always come from Variant A and the JSR dataset, not from this stylised grain.

Together, these files form a reproducible bridge between the raw JSR launch log and the final visual story used for the Maven Space Challenge poster.

If you clone this repository, place the CSV files in `data/`, and knit this R Markdown file, you should see the same KPIs and PNGs.  
If anything drifts from the numbers reported on the poster (for example, 6,617 launches, 6,126 successes, 408–410 failures, 81 partial/other), please open an issue on GitHub so we can trace and document the change.

---

*Every launch here is both a line of code and a line in history.  
My hope is that this repo makes it easy for others to audit, remix, and extend the Liftoff Atlas so the record of humanity’s climb into orbit stays transparent and open.*
