---
title: "Liftoff Atlas | JSR Data Engineering Pipeline"
subtitle: "From JSR launchlog.tsv (27k rows) 6,617 unique orbital launches"
author: "Dev Verma (@dataninja88)"
date: "`r Sys.Date()`"
output:
html_document:
---

|This document reconstructs the full data engineering workflow used to
derive a canonical set of 6,617 unique orbital launches (1957–2024)
from the Jonathan’s Space Report (JSR) launchlog.tsv file. The pipeline
prioritizes reproducibility, transparent assumptions, and minimal
transformation logic for peer validation.|

## Loading Packages

```{r}
knitr::opts_chunk$set(
echo = TRUE,
warning = FALSE,
message = FALSE
)

library(readr)
library(dplyr)
library(lubridate)
library(janitor)
library(stringr)
```

## Step 1: Load Raw JSR Launch Log

The original Jonathan’s Space Report launch log is stored as a TSV file,
data/launchlog.tsv, with one row per payload or object.

```{r}

jsr_raw <- read_tsv(
  "data/launchlog.tsv",
  col_names = c(
    "number_launch_tag",
    "launch_date",
    "piece",
    "type",
    "name",
    "pl_name",
    "jcat",
    "sat_owner",
    "sat_state",
    "lv_type",
    "flight_id",
    "platform",
    "launch_site",
    "launch_pad",
    "ascent_site",
    "ascent_pad",
    "agency",
    "lv_state",
    "launch_code",
    "lt_cite"
  ),
  trim_ws = TRUE,
  show_col_types = FALSE
)
```

## Step 2 : Clean Column Names and Parse Dates

JSR uses several date formats, for example:

"1957 Oct 4 1928:34"
"1957 Oct 4 19:28"
"1957 Oct 4"

I defined a small helper to parse these safely, then:

Create a POSIXct datetime dt.
Extract year.
Keep only launches up to and including 2024.

```{r}

jsr_raw <- jsr_raw |>
clean_names()

parse_jsr_dt <- function(x){
x <- str_squish(x)
suppressWarnings(
parse_date_time(
x,
orders = c(
"Y b d HMS",
"Y b d HM",
"Y b d H",
"Y b d",
"Y-m-d HMS",
"Y-m-d"
)
)
)
}

jsr1 <- jsr_raw |>
mutate(
dt   = parse_jsr_dt(launch_date),
year = year(dt)
) |>
filter(!is.na(dt), year <= 2024)

cat("After date parsing + year filter:", nrow(jsr1), "rows\n")
summary(jsr1$year)
```

## Step 3: Normalise Key Text Fields
For robust grouping and ID construction, I normalised:

rocket from lv_type
agency from agency
location from launch_site
pad_raw from launch_pad
flight_id_clean from flight_id

```{r}
normalize <- function(x){
x |>
str_squish() |>
str_to_lower()
}

jsr2 <- jsr1 |>
mutate(
rocket          = normalize(lv_type),
agency          = normalize(agency),
location        = normalize(launch_site),
pad_raw         = normalize(launch_pad),
flight_id_clean = normalize(flight_id)
)

cat("After normalisation:", nrow(jsr2), "rows\n")
dplyr::glimpse(jsr2)
```

## Step 4: Filter to Orbital / Failure Launch Codes
JSR’s launch_code encodes outcome and type.
Here I keep launches whose codes start with O or F which match your original filter and downstream counts.

```{r}
jsr_orbital <- jsr2 |>
filter(str_detect(launch_code, "^[OF]"))

cat("After orbital / failure filter:", nrow(jsr_orbital), "rows\n")
head(table(jsr_orbital$launch_code), 10)
```

## Step 5: Collapse Payload Rows → One Row per Launch
JSR is still payload-level.
To get one row per launch, we build a stable launch_id using:

launch time rounded to the minute
location
pad_raw
rocket
flight_id_clean

That combination is what collapses the 23k+ payload rows to 6,617 unique launches.

```{r}
jsr_launches <- jsr_orbital |>
mutate(
dt_min = floor_date(dt, unit = "minute"),
launch_id = str_c(
format(dt_min, "%Y-%m-%d %H:%M"),
location,
pad_raw,
rocket,
flight_id_clean,
sep = " | "
)
) |>
group_by(launch_id) |>
summarise(
dt_min         = min(dt_min, na.rm = TRUE),
site_raw       = first(location),
pad_raw        = first(pad_raw),
vehicle        = first(rocket),
agency         = first(agency),
lv_state       = first(lv_state),
launch_code    = first(launch_code),
n_payload_rows = n(),
.groups        = "drop"
) |>
arrange(dt_min)

cat("Unique launches:", nrow(jsr_launches), "\n")
length(unique(jsr_launches$launch_id))
head(jsr_launches$launch_id, 3)
```

## Step 6: Classify Outcomes and Save Canonical CSV

Now map launch_code into simple outcome labels and save the dataset

```{r}
jsr_launches <- jsr_launches |>
mutate(
outcome = case_when(
str_starts(launch_code, "OS") ~ "Success",
str_starts(launch_code, "OF") ~ "Failure",
str_starts(launch_code, "OP") |
str_starts(launch_code, "OU") ~ "Partial / other",
TRUE ~ "Other"
),
success_hard = outcome == "Success",
success      = success_hard
)

jsr_launches |>
count(outcome, name = "launches") |>
mutate(
total = sum(launches),
share = round(100 * launches / total, 1)
)

# Save rebuilt dataset (R project `data/` folder)

write_csv(jsr_launches, "data/jsr_unique_launches_1957_2024_rebuilt.csv")

cat("Saved: data/jsr_unique_launches_1957_2024_rebuilt.csv\n")
```


End Notes|

The pipeline reduces the original ~27,239 JSR records to 6,617 unique
orbital launches after rigorous cleaning, normalization, timestamp
alignment, and payload-row collapsing. This dataset now serves as the
canonical source for all downstream Liftoff Atlas visualizations:
timeline, world map, site rankings, vehicle trends, and Cosmic Swarm.